---
title: "Rhesus and Human Single Cell Slingshot"
output: html_document
date: "2025-06-26"
---
---
title: "Slingshot Human vs Rhesus"
output: html_document
date: "2024-04-11"
---
```{r}
library(Seurat)
library(ggplot2)
library(sctransform)
library(SeuratWrappers)
library(cellranger)
library(monocle3)
library(magrittr)
library(data.table)
library(slingshot)
library(clustree)
library(patchwork)
library(dplyr)
library(ggplot2)
library(viridis)
library(RColorBrewer)
library(GGally)
library(gplots)
library(plotly)
library(scales)
library(SingleCellExperiment)
library(ggbeeswarm)
library(CellChat)
library(tradeSeq)
library(ComplexHeatmap)
library(colorRamp2)
```



```{r}
##Load seurat objects

hCOTO <- readRDS("/data/hCOTO.harm.FINAL.2.14.24.rds")
rCOTO <- readRDS("/data/rCOTO.harm.FINAL.2.2.24.rds")
rhCOTO <- readRDS("/data/HumanvsRhesus.harm.2.14.24.rds")
rhCOTO_rhesus <- readRDS("/data/HumanvsRhesus.harm_rhesus.6.19.24.rds")
rhCOTO_human <- readRDS("/data/HumanvsRhesus.harm_human.6.19.24.rds")
```

```{r}
DimPlot(rCOTO, reduction = "umap", label.size = 6,pt.size = 2, label=T, cols = c("CTB-1"= "#1e1eff", "CTB-2"="#75abef", "CTBp-1"= "#FF7300", "CTBp-2"= "#FFE926", "STB"= "#F64DFF"))
DimPlot(rCOTO, reduction = "umap", label.size = 6,pt.size = 2, label=T, cols = c("CTB-1"= "#1e1eff", "CTB-2"="#75abef", "CTBp-1"= "#FF7300", "CTBp-2"= "#FFE926", "STB"= "#F64DFF"), split.by="orig.ident")

DimPlot(hCOTO, reduction = "umap", label.size = 6,pt.size = 2, label=T, cols = c("CTBp-1"= "#1e1eff", "CTBp-2"="#75abef", "CTB"= "#FF7300", "EVT"= "#FFE926", "STB"= "#F64DFF"))
DimPlot(hCOTO, reduction = "umap", label.size = 6,pt.size = 2, label=T, cols = c("CTBp-1"= "#1e1eff", "CTBp-2"="#75abef", "CTB"= "#FF7300", "EVT"= "#FFE926", "STB"= "#F64DFF"), split.by="orig.ident")

DimPlot(rhCOTO_rhesus, reduction = "umap", label.size = 6,pt.size = 2, label=T, cols = c("CTBp-1"= "#1e1eff", "CTBp-2"="#75abef", "CTB-1"= "#FF7300", "CTB-2"= "#FFE926", "STB"= "#F64DFF"), split.by="orig.ident")


DimPlot(rhCOTO_human, reduction = "umap", label.size = 6,pt.size = 2, label=T, cols = c("CTBp-1"= "#1e1eff", "CTBp-2"="#75abef", "CTB-1"= "#FF7300", "CTB-2"= "#FFE926", "STB"= "#F64DFF"), split.by="orig.ident")
```


```{r}
DefaultAssay(hCOTO)  <- "SCT"
DefaultAssay(rCOTO)  <- "SCT"
DefaultAssay(rhCOTO_human) <- "SCT"
DefaultAssay(rhCOTO_rhesus) <- "SCT"


hCOTOtraj <- subset(hCOTO, idents = c("CTBp-1", "CTBp-2", "CTB", "STB"))
hCOTO.sce <- as.SingleCellExperiment(hCOTOtraj)
rCOTO.sce <- as.SingleCellExperiment(rCOTO)
rhCOTO_human.sce <- as.SingleCellExperiment(rhCOTO_human)
rhCOTO_rhesus.sce <- as.SingleCellExperiment(rhCOTO_rhesus)
```

```{r}
#Run Slingshot
hCOTO.scecl <- slingshot(hCOTO.sce, clusterLabels = 'ident', reducedDim = "PCA",allow.breaks = FALSE)

rCOTO.scecl <- slingshot(rCOTO.sce, clusterLabels = 'ident', reducedDim = "PCA",allow.breaks = FALSE)

rhCOTO_human.scecl <- slingshot(rhCOTO_human.sce, clusterLabels = 'ident', reducedDim = "PCA",allow.breaks = FALSE)

rhCOTO_rhesus.scecl <- slingshot(rhCOTO_rhesus.sce, clusterLabels = 'ident', reducedDim = "PCA",allow.breaks = FALSE)
```

```{r}
slo1 <- SlingshotDataSet(hCOTO.scecl)
slo1
slo2 <- SlingshotDataSet(rCOTO.scecl)
slo2
slo3 <- SlingshotDataSet(rhCOTO_human.scecl)
slo3
slo4 <- SlingshotDataSet(rhCOTO_rhesus.scecl)
slo4
```


```{r}
DefaultAssay(rhCOTO)  <- "SCT"
noCTB2 <- subset(rhCOTO, idents = c("CTBp-1", "CTBp-2", "CTB-1", "STB"))
rhCOTO.sce <- as.SingleCellExperiment(noCTB2)
```

```{r}
rhCOTO.scecl <- slingshot(rhCOTO.sce, clusterLabels = 'ident', reducedDim = "PCA",allow.breaks = FALSE)
```

```{r}
slo5 <- SlingshotDataSet(rhCOTO.scecl)
slo5


```{r}
counts3 <- as.matrix(noCTB2@assays$RNA@counts)
filt_counts3 <- counts3[rowSums(counts3 > 5) > ncol(counts3)/100, ]
BPPARAM <- BiocParallel::bpparam()
BPPARAM$workers <- 7
sce5 <- fitGAM(counts = filt_counts3, sds=slo3,
               nknots = 5, verbose = TRUE, conditions =as.factor(noCTB2$stim))
```

```{r}
assoRes_humanvsrhesusTO <- associationTest(sce3, lineages = TRUE)
write.csv(assoRes_humanvsrhesusTO, file="assoRes_humanvsrhesusTO_lower knots.4.20.24.csv")
```

```{r}
saveRDS(sce5, file = "hversusrCOTO_fitGAM_4.21.24.rds")
```

```

```{r}
#Run Fitgam
counts <- as.matrix(hCOTOtraj@assays$RNA@counts)
filt_counts <- counts[rowSums(counts > 5) > ncol(counts)/100, ]
BPPARAM <- BiocParallel::bpparam()
BPPARAM$workers <- 7
sce <- fitGAM(filt_counts,
                       sds=slo1,
                       nknots = 4, sce = TRUE, verbose=TRUE)

saveRDS(sce, file = "hCOTO_fitGAM_4.19.24.rds")

counts2 <- as.matrix(rCOTO@assays$RNA@counts)
filt_counts2 <- counts2[rowSums(counts2 > 5) > ncol(counts2)/100, ]
BPPARAM <- BiocParallel::bpparam()
BPPARAM$workers <- 7
sce2 <- fitGAM(filt_counts2,
                       sds=slo2,
                       nknots = 4, sce = TRUE, verbose=TRUE)

saveRDS(sce2, file = "rCOTO_fitGAM_6.17.24.rds")

counts3 <- as.matrix(rhCOTO_human@assays$RNA@counts)
filt_counts3 <- counts3[rowSums(counts3 > 5) > ncol(counts3)/100, ]
BPPARAM <- BiocParallel::bpparam()
BPPARAM$workers <- 7
sce3 <- fitGAM(filt_counts3,
                       sds=slo3,
                       nknots = 4, sce = TRUE, verbose=TRUE)
saveRDS(sce3, file = "rhCOTO_human_fitGAM_6.20.24.rds")


counts4 <- as.matrix(rhCOTO_rhesus@assays$RNA@counts)
filt_counts4 <- counts4[rowSums(counts4 > 5) > ncol(counts4)/100, ]
BPPARAM <- BiocParallel::bpparam()
BPPARAM$workers <- 7
sce4 <- fitGAM(filt_counts4,
                       sds=slo4,
                       nknots = 4, sce = TRUE, verbose=TRUE)
saveRDS(sce4, file = "rhCOTO_rhesus_fitGAM_6.20.24.rds")
```



```{r}
#Association Test
assoRes_hCOTO <- associationTest(sce, lineages = TRUE)
write.csv(assoRes_hCOTO, file="assoRes_hCOTO.4.19.24.csv")

assoRes_rCOTO <- associationTest(sce2, lineages = TRUE)
write.csv(assoRes_rCOTO, file="assoRes_rCOTO.4.19.24.csv")

p.adjusted <- p.adjust(assoRes_hCOTO$pvalue, method = "fdr")
assoRes_hCOTO$adjustedPValue <- p.adjusted
significantGenes <- assoRes_hCOTO[which(assoRes_hCOTO$adjustedPValue < 0.05), ]
write.csv(significantGenes, file="assoRes_hCOTO.significantGenes.4.19.24.csv")

p.adjusted <- p.adjust(assoRes_rCOTO$pvalue, method = "fdr")
assoRes_rCOTO$adjustedPValue <- p.adjusted
rhesussignificantGenes <- assoRes_rCOTO[which(assoRes_rCOTO$adjustedPValue < 0.05), ]
write.csv(significantGenes, file="assoRes_rCOTO.significantGenes.4.19.24.csv")
```


```{r}
startRes <- startVsEndTest(sce,l2fc = log2(1))

startRes <- startVsEndTest(sce2,l2fc = log2(1))
```


```{r}
order <- order(startRes$waldStat, decreasing = TRUE)
startRes100 <- names(sce)[order[1:100]]
logcounts_matrix <- assays(hCOTO.scecl)$logcounts
specific_logcounts <- logcounts_matrix[startRes100, ]
specific_logcounts_matrix <- as.matrix(specific_logcounts)

```

```{r}
order <- order(startRes$waldStat, decreasing = TRUE)
startRes200 <- names(sce2)[order[1:200]]
logcounts_matrix <- assays(rCOTO.scecl)$logcounts
specific_logcounts <- logcounts_matrix[startRes200, ]
specific_logcounts_matrix <- as.matrix(specific_logcounts)
```


```{r}
to_plot <- specific_logcounts_matrix
ptime <- hCOTO.scecl$slingPseudotime_1
ptime_order <- colnames(to_plot)[order(ptime)]
lineage_cells <- colnames(hCOTO.scecl)[!is.na(ptime)]


annotations <- colData(hCOTO.scecl)[lineage_cells, 
                            c("slingPseudotime_1", 
                              "ident")] %>% 
    as.data.frame()

ha <- HeatmapAnnotation(df = annotations, col=list(ident=c("CTBp-1"= "#1e1eff", "CTBp-2"="#75abef", "CTB"= "#FF7300", "STB"= "#F64DFF"), show_legend = TRUE))

```

```{r}
to_plot <- specific_logcounts_matrix
ptime <- rCOTO.scecl$slingPseudotime_1
ptime_order <- colnames(to_plot)[order(ptime)]
lineage_cells <- colnames(rCOTO.scecl)[!is.na(ptime)]


annotations <- colData(rCOTO.scecl)[lineage_cells, 
                            c("slingPseudotime_1", 
                              "ident")] %>% 
    as.data.frame()

ha <- HeatmapAnnotation(df = annotations, col=list(ident=c("CTB-1"= "#1e1eff", "CTB-2"="#75abef", "CTBp-1"= "#FF7300", "CTBp-2"= "#FFE926", "STB"= "#F64DFF"), show_legend = TRUE))
```


```{r}
library(colorRamp2)
Heatmap(to_plot,
         column_order = ptime_order,
         show_column_names = FALSE,
         show_row_names = TRUE, row_names_gp = gpar(fontsize = 5),
         top_annotation = ha, na_col="#0505A6",  use_raster = TRUE, raster_by_magick = TRUE, raster_magick_filter= "Bessel", col = colorRamp2(c(0, 0.5, 0.75, 1, 1.5, 2, 2.5), c("#0505A6",  "#00FFFF", "#02F002","#ADFF2F", "#FFFF00","#FFA500", "#FF0000")))
```


****Load specific file list****
```{r}
data <- read.delim("/data/Human_traj.txt")
index_order <- order(data$waldStat, decreasing = TRUE)
data <- data[index_order, ]
top_50_genes <- head(data$X, 50)
logcounts_matrix <- assays(hCOTO.scecl)$logcounts
specific_logcounts <- logcounts_matrix[top_50_genes, ]
specific_logcounts_matrix <- as.matrix(specific_logcounts)
```

```{r}
data <- read.csv("/data/rCOTO_Toptraj.csv")
index_order <- order(data$waldStat, decreasing = TRUE)
data <- data[index_order, ]
top_50_genes <- head(data$X, 50)
logcounts_matrix <- assays(rCOTO.scecl)$logcounts
specific_logcounts <- logcounts_matrix[top_50_genes, ]
specific_logcounts_matrix <- as.matrix(specific_logcounts)
```


```{r}
to_plot <- specific_logcounts_matrix
ptime <- hCOTO.scecl$slingPseudotime_1
ptime_order <- colnames(to_plot)[order(ptime)]
lineage_cells <- colnames(hCOTO.scecl)[!is.na(ptime)]
```

```{r}
to_plot <- specific_logcounts_matrix
ptime <- rCOTO.scecl$slingPseudotime_1
ptime_order <- colnames(to_plot)[order(ptime)]
lineage_cells <- colnames(rCOTO.scecl)[!is.na(ptime)]
```


```{r}
annotations <- colData(hCOTO.scecl)[lineage_cells, 
                            c("slingPseudotime_1", 
                              "ident")] %>% 
    as.data.frame()

ha <- HeatmapAnnotation(df = annotations, col=list(ident=c("CTBp-1"= "#1e1eff", "CTBp-2"="#75abef", "CTB"= "#FF7300", "STB"= "#F64DFF"), show_legend = TRUE))

```

```{r}
annotations <- colData(rCOTO.scecl)[lineage_cells, 
                            c("slingPseudotime_1", 
                              "ident")] %>% 
    as.data.frame()

ha <- HeatmapAnnotation(df = annotations, col=list(ident=c("CTB-1"= "#1e1eff", "CTB-2"="#75abef", "CTBp-1"= "#FF7300", "CTBp-2"= "#FFE926", "STB"= "#F64DFF"), show_legend = TRUE))
```


```{r}
library(colorRamp2)
Heatmap(to_plot,
         column_order = ptime_order,
         show_column_names = FALSE,
         show_row_names = TRUE, row_names_gp = gpar(fontsize = 5),
         top_annotation = ha, na_col="#0505A6",  use_raster = TRUE, raster_by_magick = TRUE, raster_magick_filter= "Bessel", col = colorRamp2(c(0, 0.5, 0.75, 1, 1.5, 2, 2.5), c("#0505A6",  "#00FFFF", "#02F002","#ADFF2F", "#FFFF00","#FFA500", "#FF0000")))
```



```{r}
DefaultAssay(rhCOTO)  <- "SCT"
noCTB2 <- subset(rhCOTO, idents = c("CTBp-1", "CTBp-2", "CTB-1", "STB"))
rhCOTO.sce <- as.SingleCellExperiment(noCTB2)
```


```{r}
data <- read.csv("/Users/cbc54/Desktop/Human_traj.csv")
index_order <- order(data$waldStat, decreasing = TRUE)
data <- data[index_order, ]
top_150_genes <- head(data$X, 150)
logcounts_matrix <- assays(rhCOTO.scecl)$logcounts
specific_logcounts <- logcounts_matrix[top_150_genes, ]
specific_logcounts_matrix <- as.matrix(specific_logcounts)
```

```{r}
data <- read.csv("/Users/arc107/rhCOTO_traj20pick_6.28.csv")
index_order <- order(data$waldStat, decreasing = TRUE)
data <- data[index_order, ]
top_200_genes <- head(data$X, 200)
logcounts_matrix <- assays(rhCOTO.scecl)$logcounts
specific_logcounts <- logcounts_matrix[top_200_genes, ]
specific_logcounts_matrix <- as.matrix(specific_logcounts)
```


```{r}
to_plot <- specific_logcounts_matrix
ptime <- rhCOTO.scecl$slingPseudotime_1
ptime_order <- colnames(to_plot)[order(ptime)]
lineage_cells <- colnames(rhCOTO.scecl)[!is.na(ptime)]
```


```{r}
#Get positive and negative genes for trajectory
library(predictSmooth)
sce <- readRDS("/data/hCOTO_fitGAM_5.1.24.rds")

smoothers <- predictSmooth(sce2, gene = rownames(sce2), nPoints = 100)
direction_df <- smoothers %>%
     group_by(gene, lineage) %>%
     arrange(time) %>%
     summarize(direction = case_when(
         all(diff(yhat) > 0) ~ "increasing",
         all(diff(yhat) < 0) ~ "decreasing",
         TRUE ~ "mixed"
     )) %>%
     ungroup()

increasing_genes <- direction_df %>% filter(direction == "increasing")
decreasing_genes <- direction_df %>% filter(direction == "decreasing")
write.csv(increasing_genes, "increasing_genes.hTO.csv", row.names = FALSE)
write.csv(decreasing_genes, "decreasing_genes.hTO.csv", row.names = FALSE)
```



```{r}
annotations <- colData(rhCOTO.scecl)[lineage_cells, 
                                    c("slingPseudotime_1", 
                                      "ident")] %>% 
    as.data.frame()

ha <- HeatmapAnnotation(df = annotations, col=list(ident=c("CTBp-1"= "#1e1eff", "CTBp-2"="#75abef", "CTB-1"= "#FF7300", "STB"= "#F64DFF"), show_legend = TRUE))
```


```{r}
#Get positive and negative genes for trajectory
library(predictSmooth)
sce <- readRDS("/data/hCOTO_fitGAM_5.1.24.rds")

smoothers <- predictSmooth(sce2, gene = rownames(sce2), nPoints = 100)
direction_df <- smoothers %>%
     group_by(gene, lineage) %>%
     arrange(time) %>%
     summarize(direction = case_when(
         all(diff(yhat) > 0) ~ "increasing",
         all(diff(yhat) < 0) ~ "decreasing",
         TRUE ~ "mixed"
     )) %>%
     ungroup()

increasing_genes <- direction_df %>% filter(direction == "increasing")
decreasing_genes <- direction_df %>% filter(direction == "decreasing")
write.csv(increasing_genes, "increasing_genes.hTO.csv", row.names = FALSE)
write.csv(decreasing_genes, "decreasing_genes.hTO.csv", row.names = FALSE)
```



```{r}
annotations <- colData(hCOTO.scecl)[lineage_cells, 
                                    c("slingPseudotime_1", 
                                      "ident")] %>% 
    as.data.frame()

ha <- HeatmapAnnotation(df = annotations, col=list(ident=c("CTBp-1"= "#1e1eff", "CTBp-2"="#75abef", "CTB"= "#FF7300", "STB"= "#F64DFF"), col=list(slingPseudotime_1=c("#FFFFFF", "#0505A6")), show_legend = TRUE))

pdf("hTOpseudo150.pdf")
hTOpseudo150 <- Heatmap(to_plot,
        column_order = ptime_order,
        show_column_names = FALSE,
        show_row_names = TRUE, row_names_gp = gpar(fontsize = 5),
        top_annotation = ha, na_col="#0505A6",  use_raster = TRUE, raster_by_magick = TRUE, raster_magick_filter= "Bessel", col = colorRamp2(c(0, 0.5, 0.75, 1, 1.5, 2, 2.5), c("#0505A6",  "#00FFFF", "#02F002","#ADFF2F", "#FFFF00","#FFA500", "#FF0000")))
dev.off()
```



```



